Ensemble Transport Adaptive Importance Sampling

### 2019-01-10 ETAIS.tex
Made changes suggested by Colin
Changes to abstract, intro and conclusions. AMR -> MT

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and
  complexity, the need for efficient methods which can exploit the
  parallel architectures which are prevalent in high performance
  computing increases. In this paper, we aim to develop a framework
  for scalable parallel sampling algorithms. At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms naive
  parallelisation of serial MCMC methods for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\red
    for
  which speed-ups of orders of magnitude are demonstrated.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  parallel.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble. The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. We also detail how algorithmic
parameters can be quickly tuned into optimal regimes, with respect to
the effective sample size statistic. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models. The method that we present is specifically designed to
tackle the challenges of certain types of low-dimensional inverse
problem which have complex posterior structure, for example
multimodality or strong correlations, and expensive
likelihood evaluations. We will demonstrate that this methodology
allows us to sample from such a posterior distribution with greater
precision per likelihood evaluation than an uncorrelated ensemble of
standard MCMC chains of the same type.

@@

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:ETAIS}
we present the general framework of the ETAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of ETAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:AMR} we introduce the approximate
multinomial resampling (AMR) algorithm which is a less accurate but
faster alternative to resamplers which solve the optimal transport
problem exactly. In Section~\ref{sec:consistency} we consider
consistency of the ETAIS algorithm. {\red In Section~\ref{sec:useful} we briefly look at one
particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
some numerical examples, before a brief conclusion and discussion in
Section \ref{Sec:Conc}.

@@

These deficiencies of the naive method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(ETAIS). In the next section we will introduce the method in its most
general form.

@@

Initially we
will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:AMR} where a
large ensemble is required, and for which the ETPF may become more expensive. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:ETAIS}.

@@

\section{Approximate Multinomial Resampling}\label{sec:AMR}

@@

Ideally, we would like to use a resampling algorithm which is not
prohibitively costly for moderately or large sized ensembles,
which preserves the mean of the samples, and which makes it much
harder for the new samples to forget a significant region in the
density. This motivates the following algorithm, which we refer to as the
approximate multinomial resampler (AMR), which is a greedy
approximation of the ETPF resampler.

Instead of sampling $M$ times from an $M$-dimensional multinomial
distribution as is the case with the bootstrap algorithm, we sample
once each from $M$ different multinomials. Suppose that we have $M$
samples $y_n$ with weights $w_n$. The multinomial sampled from in the
bootstrap filter has a vector of probabilities given by:
\begin{equation*}
\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_M] = \bar{\bf w},
\end{equation*}
with associated states $y_n$.
We wish to find $M$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_M\}
\subset \mathbb{R}^M_{\ge 0}$
such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
given by a sample from each of the multinomials defined by the vectors
${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf y}_i$. Alternatively, as with the ETPF, a deterministic sample
can be chosen by picking each sample to be equal to the mean value of
each of these multinomial distributions, i.e. each new sample
$\hat{x}_i$ is given by:
\begin{equation}
\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,M\}.
\end{equation}

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The approximate multinomial resampler (AMR).\label{alg:AMR}}
\end{algorithm}
\end{table}

Algorithm~\ref{alg:AMR} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials as the new samples. This resampler was designed with the aims of being
numerically cheaper than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

\begin{figure}[htb]
\centering
\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX"}}
\subfigure[Relative error in
$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX2"}}\\
\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
\caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
\label{fig:AMR}
\end{figure}

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
resamplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods, where timings have been taken from simulations on a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

We will use the AMR in the numerics in Section \ref{sec:mixture}
where we have chosen to use a larger ensemble size. We do not claim
that the AMR is the optimal choice within ETAIS, but it does have
favourable features, and demonstrates how different choices of
resampler can affect the speed and efficiency of the ETAIS algorithm.

@@

@@

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes particles to new modes as they are found. This means
that we expect the number of particles in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\red In particular, we will look at the effect
of using either the ETPF, AMR or a standard bootstrap resampler.

@@

\caption{Left: Bimodal target density. Right: Convergence of RWMH and
  ETAIS-RW with ETPF, AMR and bootstrap resamplers. }
\label{kshdbf}
\end{figure}

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, AMR and bootstrap
resamplers. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the AMR is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:AMR}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates the advantage that ensemble-based methods can have over
other methods when the target
is multimodal.

@@

In the numerics which follow we have increased the ensemble size from
$M=50$ to $M=500$ to compensate for the increase in dimension. We also
use the AMR algorithm for the resampling step because of the reduced
computational cost. The method otherwise remains the same as in
previous examples. We perform test runs to find the optimal scaling
parameters considering convergence to modes with equal density. We
then calculate the convergence rates of the algorithms by producing 10
million samples from the posterior with each algorithm, and repeat the
simulation 32 times.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
\caption{Convergence of the ETAIS algorithm for the mixture model described in Section~\ref{sec:mixture}, convergence judged using the criterion in Equation~\eqref{eqn:mode_prop}. Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}. Resampling is performed using the AMR scheme.}
\label{fig:mixture_modes}
\end{figure}

@@

For this example, we use an ensemble size of {\red $M=1500$}, and the AMR resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (AMR). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Here the ETAIS results were produced with a serial
implementation, so this cost may be slightly overoptimistic, since the
runtimes do not include extra overheads of communicating the ensemble
states between processors. However, putting this together, we arrive at a speed-up factor of
over 430, which will undoubtedly overshadow any such underestimate of
the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one. 
}

@@

We have explored the application of parallelised MCMC algorithms in
low dimensional inverse problems. We have demonstrated numerically
that these algorithms converge faster than the analogous naively parallelised
Metropolis-Hastings algorithms. Further experimentation with the Metropolis
Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l|l|}
\hline
Acronym & Full name                              \\ \hline
MCMC    & Markov chain Monte Carlo               \\
RWMH    & Random walk Metropolis-Hastings        \\
MALA    & Metropolis-adjusted Langevin algorithm \\
AIS     & Adaptive importance sampling           \\
PMC     & Population Monte Carlo                 \\
ETAIS    & Ensemble transport adaptive importance sampling  \\
ETPF    & Ensemble transport particle filter     \\
AMR     & Approximate multinomial resampler      \\
ETAIS-X  & ETAIS with X kernels                    \\
AETAIS-X & Adaptive ETAIS with X kernels          \\ \hline
\end{tabular}
\end{table}

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. {\red As applications of these methods increase in size and
  complexity, the need for efficient methods increases. In this
paper, we present a particle ensemble algorithm.} At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms MCMC methods
  {\red with equivalent proposal distributions} for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, {\red multinomial transformation (MT),} which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\red
    for
  which speed-ups of orders of magnitude are demonstrated. Likelihood
  evaluations of the ensemble could be computed in a distributed
  manner, suggesting that this methodology could be a good candidate
  for parallel Bayesian computations.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  {\red ensemble, resampling}.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

{\red In this paper, we present a framework for ensemble
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble.} The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. We also detail how algorithmic
parameters can be quickly tuned into optimal regimes, with respect to
the effective sample size statistic. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models. The method that we present is specifically designed to
tackle the challenges of certain types of low-dimensional inverse
problem which have complex posterior structure, for example
multimodality or strong correlations, and expensive
likelihood evaluations. We will demonstrate that this methodology
allows us to sample from such a posterior distribution with greater
precision per likelihood evaluation than 
standard MCMC chains of the same type.

@@

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:ETAIS}
we present the general framework of the ETAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of ETAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:MT} we introduce the {\red multinomial transformation (MT)} algorithm which is a less accurate but
faster alternative to resamplers which solve the optimal transport
problem exactly. In Section~\ref{sec:consistency} we consider
consistency of the ETAIS algorithm. {\red In Section~\ref{sec:useful} we briefly look at one
particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
some numerical examples, before a brief conclusion and discussion in
Section \ref{Sec:Conc}.

@@

{\red These deficiencies of MCMC methods
motivated the development of the Ensemble Transport Adaptive Importance Sampler
(ETAIS).}
In the next section we will introduce the method in its most
general form.

@@

Initially we
will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:MT} where a
large ensemble is required, and for which the ETPF may become more expensive. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:ETAIS}.

@@

\section{Approximate Multinomial Resampling}\label{sec:MT}

@@

Ideally, we would like to use a resampling algorithm which is not
prohibitively costly for moderately or large sized ensembles,
which preserves the mean of the samples, and which makes it much
harder for the new samples to forget a significant region in the
density. This motivates the following algorithm, which we refer to as the
multinomial transformation (MT), which is a greedy
approximation of the ETPF resampler.

Instead of sampling $M$ times from an $M$-dimensional multinomial
distribution as is the case with the bootstrap algorithm, we sample
once each from $M$ different multinomials. Suppose that we have $M$
samples $y_n$ with weights $w_n$. The multinomial sampled from in the
bootstrap filter has a vector of probabilities given by:
\begin{equation*}
\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_M] = \bar{\bf w},
\end{equation*}
with associated states $y_n$.
We wish to find $M$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_M\}
\subset \mathbb{R}^M_{\ge 0}$
such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The MT is then
given by a sample from each of the multinomials defined by the vectors
${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf y}_i$. Alternatively, as with the ETPF, a deterministic sample
can be chosen by picking each sample to be equal to the mean value of
each of these multinomial distributions, i.e. each new sample
$\hat{x}_i$ is given by:
\begin{equation}
\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,M\}.
\end{equation}

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The approximate multinomial resampler (MT).\label{alg:MT}}
\end{algorithm}
\end{table}

Algorithm~\ref{alg:MT} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials as the new samples. This resampler was designed with the aims of being
numerically cheaper than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

\begin{figure}[htb]
\centering
\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX"}}
\subfigure[Relative error in
$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX2"}}\\
\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
\caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
\label{fig:MT}
\end{figure}

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and MT), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:MT} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
resamplers. As expected, the MT lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
MT preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:MT} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods, where timings have been taken from simulations on a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the MT behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

We will use the MT in the numerics in Section \ref{sec:mixture}
where we have chosen to use a larger ensemble size. We do not claim
that the MT is the optimal choice within ETAIS, but it does have
favourable features, and demonstrates how different choices of
resampler can affect the speed and efficiency of the ETAIS algorithm.

@@

{\red The numerics contained herein compare the ETAIS approach that we
  have presented against Metropolis-Hastings methods with
  the same type of proposal as is used by the kernels within
  ETAIS. Since we propose that ETAIS is a potential framework for
  parallelised Bayesian computations, we compare ETAIS with $N$
  ensemble members against an independent ensemble of $N$
  Metropolis-Hastings chains. All implementations were
  computed in serial, but the results demonstrate a clear speed-up
  which could be further exploited through parallelisation in the future.
  }

@@

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes particles to new modes as they are found. This means
that we expect the number of particles in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\red In particular, we will look at the effect
of using either the ETPF, MT or a standard bootstrap resampler.

@@

\caption{Left: Bimodal target density. Right: Convergence of RWMH and
  ETAIS-RW with ETPF, MT and bootstrap resamplers. }
\label{kshdbf}
\end{figure}

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, MT and bootstrap
resamplers. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the MT is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:MT}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates the advantage that ensemble-based methods can have over
other methods when the target
is multimodal.

@@

In the numerics which follow we have increased the ensemble size from
$M=50$ to $M=500$ to compensate for the increase in dimension. We also
use the MT algorithm for the resampling step because of the reduced
computational cost. The method otherwise remains the same as in
previous examples. We perform test runs to find the optimal scaling
parameters considering convergence to modes with equal density. We
then calculate the convergence rates of the algorithms by producing 10
million samples from the posterior with each algorithm, and repeat the
simulation 32 times.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
\caption{Convergence of the ETAIS algorithm for the mixture model described in Section~\ref{sec:mixture}, convergence judged using the criterion in Equation~\eqref{eqn:mode_prop}. Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}. Resampling is performed using the MT scheme.}
\label{fig:mixture_modes}
\end{figure}

@@

For this example, we use an ensemble size of {\red $M=1500$}, and the MT resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (MT). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Here the ETAIS results were produced with a serial
implementation, so this cost may be slightly overoptimistic, since the
runtimes do not include extra overheads of communicating the ensemble
states between processors. However, putting this together, we arrive at a speed-up factor of
over 430, which will undoubtedly overshadow any such underestimate of
the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one. 
}

@@

{\red We have explored the application of 
low dimensional Bayesian inverse problems. We have demonstrated numerically
that this method converges faster than the analogous naively parallelised
Metropolis-Hastings algorithms.} Further experimentation with the Metropolis
Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l|l|}
\hline
Acronym & Full name                              \\ \hline
MCMC    & Markov chain Monte Carlo               \\
RWMH    & Random walk Metropolis-Hastings        \\
MALA    & Metropolis-adjusted Langevin algorithm \\
AIS     & Adaptive importance sampling           \\
PMC     & Population Monte Carlo                 \\
ETAIS    & Ensemble transport adaptive importance sampling  \\
ETPF    & Ensemble transport particle filter     \\
MT     & Multinomial transformation      \\
ETAIS-X  & ETAIS with X kernels                    \\
AETAIS-X & Adaptive ETAIS with X kernels          \\ \hline
\end{tabular}
\end{table}
